Augmentation of a Historical Harpsichord Keyboard Replica for Haptic-Enabled Interaction in Museum Exhibitions

### 2025-01-22 src/article.tex
Updates from Overleaf

Here’s the extended **Hardware Design** section in LaTeX, revised to meet a target of approximately 700 words while maintaining a coherent and logical structure:

@@

The hardware design aimed to fulfil the constraints identified earlier, including non-invasiveness, reliability, and scalability, while ensuring a faithful recreation of the tactile response and auditory characteristics of historical harpsichords. To achieve these goals, the system evolved through iterative prototyping, beginning with simple threshold-based testing and culminating in a fully functional multi-sensor interface capable of triggering MIDI events.

@@

The initial stage of development focused on testing whether sensor data could reliably trigger MIDI playback. Modifying an existing harpsichord for testing was considered but ultimately discarded due to significant discrepancies in internal measurements and layout. Instead, the team employed a custom 3-key harpsichord mechanism, crafted by Graziano Bandini and provided by San Colombano (Figure \ref{fig:3key}), as a foundation for prototyping. This approach followed a methodology similar to that used in Timmermans' Haptic Key project \cite{Timmermans2020}, enabling incremental validation of hardware and sensor integration.

@@

@@

\begin{itemize}
    \item \textbf{Non-invasive:} Sensors must not require significant modifications to the harpsichord's mechanical structure or aesthetics.
    \item \textbf{Low Latency:} Sensor response times were required to remain under 10 ms, with an upper limit of 25 ms deemed acceptable \cite{Jack2016}.
    \item \textbf{Reliability:} Sensor data had to be accurate, minimising false positives and negatives without excessive filtering.
    \item \textbf{Scalability:} The design needed to be cost-effective and adaptable for up to 50 or more keys while maintaining consistent performance.
    \item \textbf{Expandability:} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

@@

@@

The QRE1113 sensors function by detecting infrared light reflected from nearby surfaces. Gradient stickers were affixed to each jack to provide a consistent reflective surface for the sensors, enabling precise tracking of jack displacement. This approach was chosen for its simplicity and scalability, allowing the design to progress from the 3-key prototype to a 49-key model without significant modifications.

To eliminate cross-talk between adjacent sensors, 3D-printed baffles were installed on the PCBs (Figure \ref{fig:baffles}). These baffles, fabricated from dark pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-} 
  \caption{Underside of the full model keyboard, showing two chambers: the front chamber (top) and the rear chamber (bottom).} 
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

The controller board, designed around the Arduino Nano BLE platform, served as the central hub for processing sensor data and triggering MIDI messages. The Nano BLE was selected for its compact form factor, integrated BLE MIDI functionality, and native USB support. These features simplified development while ensuring compatibility with existing MIDI-based synthesis systems.

@@

Non-volatile memory, implemented using Ferroelectric RAM (FRAM), provided a reliable means of storing calibration data. This ensured that sensor thresholds and other settings could be preserved across power cycles, enhancing the system's usability in a museum setting.

@@

Calibration was a critical component of the hardware design. Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software (Figure \ref{fig:serial_monitor}).

@@

The final 49-key harpsichord replica incorporated the sensor and controller boards into a rectangular frame, deviating slightly from traditional harpsichord designs to accommodate electronic components. Ribbon cables connected the PCBs, while felt strips were interwoven between the strings to suppress vibrations and maintain the tactile response of the keys.

The assembly process, conducted at the NEMUS lab in Bologna, involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs, as well as varnishing the jacks to improve sticker adhesion. Subsequent iterations addressed these challenges, refining the system’s scalability and robustness.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{3D-printed baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/serial_monitor.png}
    \caption{Arduino IDE serial plotter used for calibration of sensor thresholds.}
    \label{fig:serial_monitor}
\end{figure}

@@

\subsection{Calibration}\label{calibration}

Finer calibration carried out with the guidance of \anon{Catalina Vincens} as
an expert harpsichord performer

\begin{itemize}
\item
  sensor surface

  \begin{itemize}
  \item
    printed gradient
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  scale
\item
  time scale
\item
  working at distance
\item
  limitations

  \begin{itemize}
  \item
    latency
  \item
    deprecation

    \begin{itemize}
    \item
      software

      \begin{itemize}
      \item
        EAGLE
      \end{itemize}
    \item
      hardware

      \begin{itemize}
      \item
        FRAM
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Velocity, aftertouch
\item
  second jack row
\item
  reducing latency
\item
  alternative sensor surfaces
\end{itemize}

@@

>>>>

@@

The hardware design aimed to fulfil the earlier constraints, including non-invasiveness, reliability, and scalability. To achieve these goals, the system evolved through iterative prototyping, beginning with simple threshold-based testing and culminating in a fully functional multi-sensor interface capable of triggering MIDI events.

@@

The initial stage of development focused on testing whether sensor data could reliably trigger MIDI playback. Modifying an existing harpsichord for testing was considered but ultimately discarded due to significant internal measurement and layout discrepancies. Instead, a custom 3-key harpsichord mechanism (Figure \ref{fig:3key}) was acquired from the museum and used as a foundation for prototyping. This approach followed a methodology similar to that used in Timmermans' Haptic Key project \cite{Timmermans2020}, enabling incremental validation of hardware and sensor integration.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/3-key-side-labelled.png}
    \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini at San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
    \label{fig:3key}
\end{figure}

@@

\begin{itemize}
    \item \emph{Non-invasive:} Sensors must not require significant modifications to the harpsichord's mechanical structure or aesthetics.
    \item \emph{Low Latency:} Sensor response times were required to remain under 10 ms, with an upper limit of 25 ms deemed acceptable \cite{Jack2016}.
    \item \emph{Reliability:} Sensor data had to be accurate, minimising false positives and negatives without excessive filtering.
    \item \emph{Scalability:} The design needed to be cost-effective and adaptable for up to 50 or more keys while maintaining consistent performance.
    \item \emph{Expandability:} The system should accommodate future functionality, such as additional MIDI parameters or data visualisation.
\end{itemize}

@@

% \begin{figure*}
%     \centering
%     \includegraphics[width=\linewidth]{src/images/bl}
%     \caption{3-Key Model Harpsichord Mechanism designed by Graziano Bandini at San Colombano, Bologna, 2023. Key components include A: Wrestplank, B: Key Lever, C: Key Cap, D: Soundboard, E: Jack.}
%     \label{fig:3key}
% \end{figure*}

@@

The QRE1113 sensors function by detecting infrared light reflected from nearby surfaces. The gradient stickers affixed to each jack provided a consistent reflective surface for the sensors, enabling precise tracking of jack displacement. This approach was chosen for its simplicity and scalability, allowing the design to progress from the 3-key prototype to a 49-key model without significant modifications.

To eliminate cross-talk between adjacent sensors, 3D-printed baffles were installed on the PCBs (Figure \ref{fig:baffles}). These baffles, fabricated from dark-pigmented PLA, ensured that infrared reflections from neighbouring jacks did not interfere with sensor readings. Testing confirmed that this measure significantly improved data reliability.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/simple-schematic-bw-.jpg} 
  \caption{Optical sensor in a simple voltage divider circuit. \texttt{V\_OUT} is routed to one of 8 channels on the CD4051BE multiplexer}
  \Description{} 
  \label{fig:simple-schematic}
\end{figure}

@@

The controller board, designed around the Arduino Nano BLE platform, was the central hub for processing sensor data and triggering MIDI messages. The Nano BLE was selected for its compact form factor, integrated BLE MIDI functionality, and native USB support. These features simplified development while ensuring compatibility with existing MIDI-based synthesis systems.

@@

Non-volatile memory, implemented using Ferroelectric RAM (FRAM), provided a reliable means of storing calibration data. This ensured that sensor thresholds and other settings could be preserved across power cycles, enhancing the system's usability in museums.

@@

Each sensor's threshold was manually adjusted using RGB LEDs integrated into the system. These LEDs provided visual feedback during calibration, enabling the identification of malfunctioning sensors and simplifying the alignment process. The calibration workflow utilised the Arduino IDE serial plotter and open-source MIDI monitoring software (Figure \ref{fig:serial_monitor}).

Finer calibration carried out with the guidance of \anon{Catalina Vincens} as an expert harpsichord performer.

@@

The final 49-key harpsichord replica incorporated the sensor and controller boards into a rectangular frame, deviating slightly from traditional designs to accommodate electronic components. Ribbon cables connected the PCBs, while felt strips were interwoven between the strings to suppress vibrations and maintain the tactile response of the keys.

The assembly process at the NEMUS lab in Bologna involved meticulous adjustments to key clearances and wiring layouts. Modifications included milling sections of the keys to ensure adequate clearance for the PCBs and varnishing the jacks to improve sticker adhesion. Subsequent iterations addressed these challenges, refining the system’s scalability and robustness.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/baffles.png}
    \caption{STL Model of baffles designed to prevent cross-talk between adjacent sensors.}
    \label{fig:baffles}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{src/images/serial_monitor.png}
    \caption{Arduino IDE serial plotter used for calibration of sensor thresholds. \texttt{M} and \texttt{m} represent the maximum an minimum value of digitised data. \texttt{K} and \texttt{T} are the values of the sensor and threshold being calibrated, followed by a two digit number denoting the sensor index (in this case sensor 32).}
    \label{fig:serial_monitor}
\end{figure}

@@

@@

Paper has presented the design for an interactive exhibition at a musical instrument museum. Design attempts to address the problem.
Optical sensors read by a microcontroller have been shown to be a solution to tracking vertical displacement of harpsichord jacks and therefore a viable system to augmenting harpsichords. The interface's capabilities of the jack rails to disengage and to convert string plucks to MIDI data means this can be used as a probe to explore the how the ``double-pluck'' can be used as vector for expression during performance. The placement of the instrument as a museum exhibit means it also has potential to be used in a long form study.

@@

The current interface is an advanced prototype, but there remain many avenues to explore and aspects of the design to improve.

@@

% \begin{tikzpicture}[auto, node distance=2cm,>=latex']

%     \node [input, name=input] {};
%     \node [sum, right of=input] (sum) {};
%     \node [block, right of=sum] (controller) {Controller};
%     \node [block, right of=controller, pin={[pinstyle]above:D},
%             node distance=3cm] (system) {System};

%     \draw [->] (controller) -- node[name=u] {$u$} (system);
%     \node [output, right of=system] (output) {};
%     \node [block, below of=u] (measurements) {Measurements};

%     \draw [draw,->] (input) -- node {$r$} (sum);
%     \draw [->] (sum) -- node {$e$} (controller);
%     \draw [->] (system) -- node [name=y] {$y$}(output);
%     \draw [->] (y) |- (measurements);
%     \draw [->] (measurements) -| node[pos=0.99] {$-$} 
%         node [near end] {$y_m$} (sum);
% \end{tikzpicture}
